---
title: "1. Installation"
description: >
  Alternatives to install rgplates. 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<br>

## Stable version 
* * *

#### From the CRAN

The stable version of the package is available on [CRAN servers](https://cran.r-project.org/package=rgplates), which means that you can install it with the regular `install.packages()` function.

```{r cran, eval=FALSE}
install.packages("rgplates")
```

If for some reason, the installlation from CRAN fails, you might need to look into alternative methods.


#### From GitHub

You can install the package from source directly from GitHub using the `[devtools](https://cran.r-project.org/package=devtools)` package. All you need is install `devtools` and then run the this line of code:

```{r gitStab, eval=FALSE}
devtools::install_github("gplates/rgplates")
```

#### From Zenodo 

For the sake of completeness and simplicity, the sources of the stable version is also deposited on Zenodo. The `install.packages()` function can be used to install this from source (although you might have install dependencies of the package. )

```{r stable, eval=FALSE}
install.packages(
	"https://zenodo.org/records/13711982/files/rgplates_0.5.0.tar.gz?download=1", 
	repos=NULL, type="source")
```

<br>

## Development version
* * * 

If you encounter technical problems with the package that are not described anywhere, you might want to take a look at the [development version](https://github.com/gplates/rgplates/tree/devel).


#### Development version from GitHub

The simplest way to install the in-development version is via the `devtools` package. You just have to indicate that you want to install from the in-development `devel` branch.

```{r gitDev, eval=FALSE}
devtools::install_github("gplates/rgplates@devel")
```

#### Manual installation

If you wish to install the development version, I recommend a manual installation:  
1. Clone the repository to your local hard drive.  
2. Open a terminal and navigate to the directory where you cloned. The `rgplates` directory should be visible from there.  
3. Run this line in the terminal  

```
R CMD INSTALL rgplates
```

- *If you see an error suggesting that `R` is not found, you have to add it to your `PATH` environmental variable.*  
- *If the R packages that `rgplates` depend on are not installed, you have to install them manually, or you will get an error.*
- *You can also use `devtools` to install the development version with:*  
